---
title: "curatedTCGAData"
date: "`r BiocStyle::doc_date()`"
vignette: |
  %\VignetteIndexEntry{curatedTCGAData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc_float: true
Package: curatedTCGAData
---

```{r, echo = FALSE, warning = FALSE}
suppressPackageStartupMessages({
    library(MultiAssayExperiment)
    library(curatedTCGAData)
    library(TCGAutils)
})
```

# Installation

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("curatedTCGAData")
```

Load packages:

```{r, eval = FALSE}
library(curatedTCGAData)
library(MultiAssayExperiment)
library(TCGAutils)
```

# Downloading datasets

Checking available cancer codes and assays in TCGA data:

```{r}
curatedTCGAData(diseaseCode = "*", assays = "*", dry.run = TRUE)
```

Check potential files to be downloaded:

```{r}
curatedTCGAData(diseaseCode = "COAD", assays = "RPPA*", dry.run = TRUE)
```

# Caveats for working with TCGA data

Not all TCGA samples are cancer, there are a mix of samples in each of the
33 cancer types. Use `sampleTables` on the `MultiAssayExperiment` object
along with `data(sampleTypes, package = "TCGAutils")` to see what samples are
present in the data. There may be tumors that were used to create multiple
contributions leading to technical replicates. These should be resolved using
the appropriate helper functions such as `mergeReplicates`. Primary tumors
should be selected using `TCGAutils::TCGAsampleSelect` and used as input
to the subsetting mechanisms. See the "Samples in Assays" section of this
vignette.

## ACC dataset example

```{r, message=FALSE}
(accmae <- curatedTCGAData("ACC", c("CN*", "Mutation"), FALSE))
```

**Note**. For more on how to use a `MultiAssayExperiment` please see the
`MultiAssayExperiment` vignette.

### Subtype information

Some cancer datasets contain associated subtype information within the
clinical datasets provided. This subtype information is included in the
metadata of `colData` of the `MultiAssayExperiment` object. To obtain these
variable names, use the `getSubtypeMap` function from TCGA utils:

```{r}
head(getSubtypeMap(accmae))
```

### Typical clinical variables

Another helper function provided by TCGAutils allows users to obtain a set
of consistent clinical variable names across several cancer types.
Use the `getClinicalNames` function to obtain a character vector of common
clinical variables such as vital status, years to birth, days to death, etc.

```{r}
head(getClinicalNames("ACC"))

colData(accmae)[, getClinicalNames("ACC")][1:5, 1:5]
```

### Samples in Assays

The `sampleTables` function gives an overview of sample types / codes
present in the data:

```{r}
sampleTables(accmae)
```

Often, an analysis is performed comparing two groups of samples to each other.
To facilitate the separation of samples, the `splitAssays` TCGAutils function
identifies all sample types in the assays and moves each into its own
assay. By default, all discoverable sample types are separated into
a separate experiment. In this case we requested only solid tumors and blood
derived normal samples as seen in the `sampleTypes` reference dataset:

```{r}
sampleTypes[sampleTypes[["Code"]] %in% c("01", "10"), ]

splitAssays(accmae, c("01", "10"))
```

To obtain a logical vector that could be used for subsetting a
`MultiAsssayExperiment`, refer to `TCGAsampleSelect`. To select only primary
tumors, use the function on the colnames of the `MultiAssayExperiment`:

```{r}
tums <- TCGAsampleSelect(colnames(accmae), "01")
```

You can subsequently provide this input to the subsetting function to select
only primary tumors:

```{r}
accmae[, tums, ]
```

# Session Information

```{r}
sessionInfo()
```

