---
title: "curatedTCGAData"
date: "`r BiocStyle::doc_date()`"
vignette: |
  %\VignetteIndexEntry{curatedTCGAData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc_float: true
Package: curatedTCGAData
---

```{r, echo = FALSE, warning = FALSE}
suppressPackageStartupMessages({
    library(MultiAssayExperiment)
    library(curatedTCGAData)
})
```

# Installation

```{r, eval=FALSE}
if (!require("BiocManager"))
    install.packages("BiocManager")

BiocManager::install("curatedTCGAData")
```

Load packages:

```{r, eval = FALSE}
library(curatedTCGAData)
library(MultiAssayExperiment)
```

# Downloading datasets

Checking available cancer codes and assays in TCGA data:

```{r}
curatedTCGAData(diseaseCode = "*", assays = "*", dry.run = TRUE)
```

Check potential files to be downloaded:

```{r}
curatedTCGAData(diseaseCode = "COAD", assays = "RPPA*", dry.run = TRUE)
```

## GBM dataset example

```{r, message=FALSE}
gbm <- curatedTCGAData("GBM", "RPPA*", FALSE)
gbm
```

**Note**. For more on how to use a `MultiAssayExperiment` please see the
`MultiAssayExperiment` vignette.

### Subtype information

Some cancer datasets contain associated subtype information within the
clinical datasets provided. This subtype information is included in the
metadata of `colData` of the `MultiAssayExperiment` object. To obtain these
variable names, run the `metadata` function on the `colData` of the object
such as:

```{r}
head(metadata(colData(gbm))[["subtypes"]])
```

